A Tale of Two Transects

### 2019-12-25 writing/transect_paper.Rmd
improve figures

---
title: "A Tale of Two Transects"
subtitle: "Also Two Sequencing Technologies, and Two Amplicons"
author: 
  - Brendan Furneaux^1^,
    Roël Houdanon^2^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
output:
  bookdown::word_document2: 
    keep_md: yes
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
---

@@

```{r region-length}
datasets <- read_csv(here("config/datasets.csv"), col_types = "cccicccccicc")
regions <- read_csv(here("config/regions.csv"), col_types = "cccciiiiic")
regions %>%
  select(Region = region, "Minimum length" = min_length, "Maximum length" = max_length) %>%
  right_join(tibble(Region = c("ITS1", "5_8S", "ITS2", "LSU1", "D1", "LSU2", "D2", "LSU3", "D3", "LSU4"))) %>%
  mutate_at("Region", str_replace_all, "_", ".") %>%
  knitr::kable(
    booktabs = TRUE,
    caption = "Minimum and maximum allowed lengths for each extracted region."
  )
```

@@

(ref:readcap) ASVs/OTUs which were detected by one sequencing strategy but not the other are plotted as tick marks along the axes. Dashed line represents a constant ratio of read numbers. The blue line is a LOESS smooth of the data, with associated uncertainty in grey shading. $R^2$ value displayed is for log-transformed non-zero read numbers.

@@

```{r pbl-vs-its, fig.cap="Comparison between read number for PacBio RSII long reads and Ion Proton S5 reads, by ASV (left) and 95% OTU (right). (ref:readcap)"}
loadd(big_seq_table_ITS2, cache = cache)
asv_table <- big_seq_table_ITS2 %>%
  tibble::as_tibble(rownames = "filename") %>%
  tidyr::gather(key = "seq", value = "reads", -1) %>%
  dplyr::filter(reads >= 1) %>%
  tidyr::extract(col = "filename",
                 into = c("seq.run", "plate", "well", "dir", "region"),
                 regex = "([a-z]+_\\d+)_(\\d+)_([A-H]1?[0-9])([fr]?)_([:alnum:]+).+") %>%
  dplyr::group_by(seq.run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq.run, value = reads, fill = 0)

@@

comparisons <- crossing(
  x = factor(datasets$seq_run, ordered = TRUE),
  y = factor(datasets$seq_run, ordered = TRUE),
  type = c("ASV", "OTU")
) %>%
  filter(y > x) %>%
  mutate_all(as.character) %>%
  mutate(
    cross = paste(x, y, sep = "-")
  ) %>%
  group_by(cross, type) %>%
  group_modify(function(d, g) {
    big_table %>%
      filter(type == g$type) %>%
      select(!!d$x, !!d$y) %>%
      filter(.data[[d$x]] > 0 & .data[[d$y]] > 0) %>%
      mutate_all(log) %>%
      lm() %>%
      broom::glance() %>%
      bind_cols(d, .)
    }) %>%
  mutate(
    label = paste("R^2 ==", formatC(r.squared, format = "f", digits = 2))
  )

@@

big_table %>%
  filter(is_057 > 0, pb_500 > 0) %>%
  ggplot(aes(x = is_057, y = pb_500)) +
  geom_point(alpha = 0.2, shape = 1) +
  scale_x_log10() +
  scale_y_log10(limits = c(1, 1e4)) +
  coord_equal() +
  xlab("Ion Torrent Short Reads") +
  ylab("PacBio Long Reads") +
  geom_rug(aes(y = pb_500, x = 1), sides = "l", alpha = 0.2,
           data = big_table %>% filter(is_057 == 0, pb_500 > 0)) +
  geom_rug(aes(y = is_057), sides = "b", alpha = 0.2,
           data = big_table %>% filter(pb_500 == 0, is_057 > 0)) +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_abline(aes(intercept = mean(log10(pb_500[pb_500 >= 1])) -
                    mean(log10(is_057[is_057 >= 1])), slope = 1),
              linetype = 2L, alpha = 0.5) +
  geom_text(
    aes(label = label),
    data = filter(comparisons, x == "is_057", y == "pb_500"),
    x = 1,
    y = 3.5,
    color = "black",
    hjust = 0,
    parse = TRUE,
    inherit.aes = FALSE
  ) +
  facet_wrap(~type)
```

- Figure \@ref(fig:pbl-vs-its)
  - ASVs/OTUs are correlated between Ion Torrent and PacBio; 
  ASVs $R^2 = `r compR2("is_057", "pb_500", "ASV")`$; 
  OTUs $R^2 = `r compR2("is_057", "pb_500", "OTU")`$
  - "hook" at lower left due to sampling effects with rare species
  - lots of ASVs/OTUs detected in only one tech

- Figure \@ref(fig:pbs-vs-its)
  - very few PacBio OTUs undetected by Ion torrent from same library; only rare OTUS undetected by PacBio
  - $R^2 = `r compR2("is_057", "pb_483", "OTU")`$
  - sequencing technologies yield comparable results (given read depth) with sufficient clustering
  - not such good match with ASVs -- suggests DADA2 doesn't perform equally on different datasets

- Figure \@ref(fig:pbl-vs-pbs)
  - No "hook", because depth is similar
  - still lots of ASVs/OTUs detected by only one amplicon -- primer bias.

```{r pbs-vs-its, message=FALSE, fig.cap="Comparison between read number for PacBio RSII short reads and Ion Proton S5 reads, by ASV (left) and 95% OTU (right). (ref:readcap)"}
big_table %>% 
  filter(is_057 > 0, pb_483 > 0) %>%
  ggplot(aes(x = is_057, y = pb_483)) +
  geom_point(alpha = 0.2, shape = 1) +
  scale_x_log10() +
  scale_y_log10(limits = c(1, 1e4)) +
  xlab("Ion Torrent Short Reads") +
  ylab("PacBio Short Reads") +
  coord_equal() +
  geom_rug(aes(y = pb_483, x = 1), sides = "l", alpha = 0.2,
           data = big_table %>% filter(is_057 == 0, pb_483 > 0)) +
  geom_rug(aes(y = is_057), sides = "b", alpha = 0.2,
           data = big_table %>% filter(pb_483 == 0, is_057 > 0)) +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_abline(aes(intercept = mean(log10(pb_483[pb_483 >= 1])) -
                    mean(log10(is_057[is_057 >= 1])), slope = 1),
              linetype = 2L, alpha = 0.5) +
  geom_text(
    aes(label = label),
    data = filter(comparisons, x == "is_057", y == "pb_483"),
    x = 1,
    y = 3.5,
    color = "black",
    hjust = 0,
    parse = TRUE,
    inherit.aes = FALSE
  ) +
  facet_wrap(~type)

```{r pbl-vs-pbs, message=FALSE, fig.cap="Comparison between read number for PacBio RSII long and short amplicon reads, by ASV (left) and 95% OTU (right). (ref:readcap)"}
big_table %>% 
  filter(pb_483 > 0, pb_500 > 0) %>%
  ggplot(aes(x = pb_483, y = pb_500)) +
  geom_point(alpha = 0.2, shape = 1) +
  scale_x_log10() +
  scale_y_log10() +
  coord_equal() +
  xlab("PacBio Short Reads") +
  ylab("PacBio Long Reads") +
  geom_rug(aes(y = pb_500, x = 1), sides = "l", alpha = 0.2,
           data = big_table %>% filter(pb_483 == 0, pb_500 > 0)) +
  geom_rug(aes(y = pb_483), sides = "b", alpha = 0.2,
           data = big_table %>% filter(pb_500 == 0, pb_483 > 0)) +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_abline(aes(intercept = mean(log10(pb_500[pb_500 >= 1])) -
                    mean(log10(pb_483[pb_483 >= 1])), slope = 1),
              linetype = 2L, alpha = 0.5) +
  geom_text(
    aes(label = label),
    data = filter(comparisons, x == "pb_483", y == "pb_500"),
    x = 1,
    y = 3.5,
    color = "black",
    hjust = 0,
    parse = TRUE,
    inherit.aes = FALSE
  ) +
  facet_wrap(~type)
```

@@

variog_empir <- variog_points %>%
  group_by(guild, metric, tech, amplicon, dist, timelag) %>%
  summarize_at("gamma", mean)

@@

variog_fit <- variog_data %>%
  mutate_at("variofitST", lapply, gstat:::vgmMetric, expand_grid(spacelag = seq(1, 25, 0.5), timelag = 0:1)) %>%
  unnest(variofitST) %>%
  rename(dist = spacelag) %>%
  mutate_at("timelag", factor)

variog_empir %>%
  ggplot(aes(dist, gamma, color = timelag)) +
  # geom_hline(data = variog_sills, aes(yintercept = sills), linetype = "dashed", color = "gray50") +
  # geom_vline(data = variog_data, aes(xintercept = range), linetype = "dashed", color = "gray50") +
  # geom_jitter(shape = 46, alpha = 0.5) +
  geom_line(size = 1.5, alpha = 0.5) +
  geom_point(size = 1.5, alpha = 0.5) +
  # geom_smooth(data = variog_points, aes(fill = timelag), color = FALSE, alpha = 0.5) +
  geom_line(data = filter(variog_fit, timelag == 1), size = 0.5, color = "darkcyan", alpha = 0.8) +
  geom_line(data = filter(variog_fit, timelag == 0), size = 0.5, color = "red1", alpha = 0.8) +
  scale_x_continuous(expand = expand_scale(0, 0), limits = c(0, 25), name = "Distance (m)") +
  scale_color_discrete() +
  scale_y_continuous(name = "Community dissimilarity") +
  facet_grid(metric + amplicon ~ guild, scales = "free")

>>>>

---
title: "A Tale of Two Transects"
subtitle: "Also Two Sequencing Technologies, and Two Amplicons"
author: 
  - Brendan Furneaux^1^,
    Roël Houdanon^2^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
output:
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
  bookdown::word_document2: 
    keep_md: yes
---

@@

```{r region-length}
datasets <- read_csv(here("config/datasets.csv"), col_types = "cccicccccicc") %>%

regions <- read_csv(here("config/regions.csv"), col_types = "cccciiiiic")
regions %>%
  select(Region = region, "Minimum length" = min_length, "Maximum length" = max_length) %>%
  right_join(tibble(Region = c("ITS1", "5_8S", "ITS2", "LSU1", "D1", "LSU2", "D2", "LSU3", "D3", "LSU4"))) %>%
  mutate_at("Region", str_replace_all, "_", ".") %>%
  knitr::kable(
    booktabs = TRUE,
    caption = "Minimum and maximum allowed lengths for each extracted region."
  )
```

@@

(ref:readcap) Comparison between read numbers for different sequencing strategies, by ASV (left) and 97% OTU (right). ASVs/OTUs which were detected by one sequencing strategy but not the other are plotted as tick marks along the axes. Dashed line represents a constant ratio of read numbers. The blue line is a LOESS smooth of the data, with associated uncertainty in grey shading. $R^2$ value displayed is for log-transformed non-zero read numbers.

@@

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
loadd(big_seq_table_ITS2, cache = cache)
asv_table <- big_seq_table_ITS2 %>%
  tibble::as_tibble(rownames = "filename") %>%
  tidyr::gather(key = "seq", value = "reads", -1) %>%
  dplyr::filter(reads >= 1) %>%
  tidyr::extract(col = "filename",
                 into = c("seq.run", "plate", "well", "dir", "region"),
                 regex = "([a-z]+_\\d+)_(\\d+)_([A-H]1?[0-9])([fr]?)_([:alnum:]+).+") %>%
  dplyr::group_by(seq.run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq.run, value = reads, fill = 0)

@@

multi_table <- crossing(
  x_var = factor(datasets$seq_run, ordered = TRUE),
  y_var = factor(datasets$seq_run, ordered = TRUE),
  type = c("ASV", "OTU")
) %>%
  filter(x_var < y_var) %>%
  mutate_all(as.character) %>%
  group_by(x_var, y_var, type) %>%
  group_modify(
    function(d, g) {
      big_table %>%
        filter(type == g$type) %>%
        select(x = !!g$x_var, y = !!g$y_var) %>%
        filter(x > 0 | y > 0)
    }
  )

comparisons <- multi_table %>%
  group_modify(
    function(d, g) {
      filter_all(d, ~. > 0) %>%
        mutate_all(log10) %>%
      {bind_cols(
      lm(y ~ x, .) %>%
        broom::glance(.),
      lm(y ~ offset(x), .) %>%
        broom::tidy(.) %>%
        select(term, estimate) %>%
        pivot_wider(names_from = term, values_from = estimate)
      )}
    }
  ) %>%
  mutate(
    label = paste("R^2 ==", formatC(r.squared, format = "f", digits = 2))
  )

# comparisons <- crossing(
#   x_var = factor(datasets$seq_run, ordered = TRUE),
#   y_var = factor(datasets$seq_run, ordered = TRUE),
#   type = c("ASV", "OTU")
# ) %>%
#   filter(y_var < x_var) %>%
#   mutate_all(as.character) %>%
#   group_by(x_var, y_var, type) %>%
#   group_modify(function(d, g) {
#     big_table %>%
#       filter(type == g$type) %>%
#       select(x = !!g$x_var, y = !!g$y_var) %>%
#       filter(.data[[g$x_var]] > 0 & .data[[g$y_var]] > 0) %>%
#       mutate_all(log10) %>%
#       lm() %>%
#       {bind_rows(
#       broom::glance(.),
#       broom::tidy(.) %>%
#         select(term, estimate) %>%
#         pivot_wider(names_from = term, values_from = estimate)
#       )} %>%
#       bind_cols(d, .)
#     }) %>%
#   mutate(
#     label = paste("R^2 ==", formatC(r.squared, format = "f", digits = 2))
#   )

@@

# p <- multi_table %>%
#   filter(x > 0, y > 0, type == "ASV") %>%
#   ggplot(aes(y = y, x = x)) +
#   geom_point(alpha = 0.2, shape = 1) +
#   scale_x_log10() +
#   scale_y_log10(limits = c(1, 1e4)) +
#   coord_equal() +
#   xlab(NULL) +
#   ylab(NULL) +
#   geom_rug(aes(y = y), sides = "l", alpha = 0.2,
#            data = multi_table %>% filter(x == 0, y > 0)) +
#   geom_rug(aes(x = x), sides = "b", alpha = 0.2,
#            data = multi_table %>% filter(y == 0, x > 0)) +
#   facet_grid(y_var ~ x_var, switch = "both", space = "free") +
#   stat_smooth(method = "loess", formula = y ~ x) +
#   geom_abline(aes(intercept = `(Intercept)`, slope = 1),
#               linetype = 2L, alpha = 0.5, data = comparisons) +
#   geom_text(
#     aes(label = label),
#     data = comparisons,
#     x = 1,
#     y = 3.5,
#     color = "black",
#     hjust = 0,
#     parse = TRUE,
#     inherit.aes = FALSE
#   ) +
#     theme(strip.placement = "outside", strip.background = element_blank())
# g <- ggplotGrob(p)
# g$grobs[g$layout$name %in% c("panel-1-2")] <- NULL
# g$layout <- g$layout[!(g$layout$name %in% c("panel-1-2")),]
# grid.newpage()
# grid.draw(g)

expand_seq_run <- function(labels) {
  lapply(
    labels,
    )
}

read_comparison <- function(multi_table, comparisons, type) {
  multi_table <- filter(multi_table, type == !!type)
  comparisons <- filter(comparisons, type == !!type)
  p <- multi_table %>%
    filter(x > 0, y > 0) %>%
    ggplot(aes(y = y, x = x)) +
    geom_point(alpha = 0.2, shape = 1) +
    scale_x_log10() +
    scale_y_log10(limits = c(1, 1e4)) +
    coord_equal() +
    xlab(NULL) +
    ylab(NULL) +
    geom_rug(aes(y = y), sides = "l", alpha = 0.2,
             data = multi_table %>% filter(x == 0, y > 0)) +
    geom_rug(aes(x = x), sides = "b", alpha = 0.2,
             data = multi_table %>% filter(y == 0, x > 0)) +
    facet_grid(y_var ~ x_var, switch = "both", space = "free", ) +
    stat_smooth(method = "loess", formula = y ~ x) +
    geom_abline(aes(intercept = `(Intercept)`, slope = 1),
                linetype = 2L, alpha = 0.5, data = comparisons) +
    geom_text(
      aes(label = label),
      data = comparisons,
      x = 0.5,
      y = 3.5,
      color = "black",
      hjust = 0,
      parse = TRUE,
      inherit.aes = FALSE
    ) +
    theme(strip.placement = "outside", strip.background = element_blank())
  g <- ggplotGrob(p)
  g$grobs[g$layout$name %in% c("panel-1-2")] <- NULL
  g$layout <- g$layout[!(g$layout$name %in% c("panel-1-2")),]
  g
}
ggpubr::ggarrange(
  read_comparison(multi_table, comparisons, type = "ASV"),
  read_comparison(multi_table, comparisons, type = "OTU"),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)

read_comparison <- function(big_table, comparisons, x, y, xlab, ylab) {
  x_var <- enquo(x)
  y_var <- enquo(y)
  big_table %>%
  filter(!!x_var > 0, !!y_var > 0) %>%
  ggplot(aes(x = !!x_var, y = !!y_var)) +
  geom_point(alpha = 0.2, shape = 1) +
  scale_x_log10() +
  scale_y_log10(limits = c(1, 1e4)) +
  coord_equal() +
  xlab(xlab) +
  ylab(ylab) +
  geom_rug(aes(y = !!y_var, x = 1), sides = "l", alpha = 0.2,
           data = big_table %>% filter(!!x_var == 0, !!y_var > 0)) +
  geom_rug(aes(y = !!x_var), sides = "b", alpha = 0.2,
           data = big_table %>% filter(!!y_var == 0, !!x_var > 0)) +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_abline(aes(intercept = mean(log10((!!y_var)[(!!y_var) >= 1])) -
                    mean(log10((!!x_var)[(!!x_var) >= 1])), slope = 1),
              linetype = 2L, alpha = 0.5) +
  geom_text(
    aes(label = label),
    data = filter(comparisons, .data$x == quo_name(x_var), .data$y == quo_name(y_var)),
    x = 1,
    y = 3.5,
    color = "black",
    hjust = 0,
    parse = TRUE,
    inherit.aes = FALSE
  ) +
  facet_wrap(~type, strip.position = "right", scales = "free")
}
# ggpubr::ggarrange(
#   ncol = 1,
#   labels = "auto",
#   read_comparison(
#     big_table,comparisons,
#     is_057, pb_500,
#     "Ion Torrent Short Reads", "PacBio Long Reads"),
#   read_comparison(
#     big_table, comparisons,
#     is_057, pb_483,
#     "Ion Torrent Short Reads", "PacBio Short Reads"),
#   read_comparison(
#     big_table, comparisons,
#     pb_483, pb_500,
#     "PacBio Short Reads", "PacBio Long Reads")
# )
```

- Figure \@ref(fig:pb-vs-it)a
  - ASVs/OTUs are correlated between Ion Torrent and PacBio; 
  ASVs $R^2 = `r compR2("is_057", "pb_500", "ASV")`$; 
  OTUs $R^2 = `r compR2("is_057", "pb_500", "OTU")`$
  - "hook" at lower left due to sampling effects with rare species
  - lots of ASVs/OTUs detected in only one tech

- Figure \@ref(fig:pb-vs-it)b
  - very few PacBio OTUs undetected by Ion torrent from same library; only rare OTUS undetected by PacBio
  - $R^2 = `r compR2("is_057", "pb_483", "OTU")`$
  - sequencing technologies yield comparable results (given read depth) with sufficient clustering
  - not such good match with ASVs -- suggests DADA2 doesn't perform equally on different datasets

- Figure \@ref(fig:pb-vs-it)c
  - No "hook", because depth is similar
  - still lots of ASVs/OTUs detected by only one amplicon -- primer bias.

@@

variog_empir <- variog_points %>%
  # mutate(bin = cut(dist, c(1:13 - 0.5, 14.5, 16.5, 19.5, 24.5))) %>%
  group_by(guild, metric, tech, amplicon, dist, timelag) %>%
  summarize(
    gamma = mean(gamma),
    # dist = mean(dist),
    n = n()
  )

@@

variog_fit <- variog_data %>%
  mutate_at("variofitST", lapply,
            gstat:::vgmMetric,
            expand_grid(spacelag = seq(1, 25, 0.5), timelag = 0:1)
            ) %>%
  unnest(variofitST) %>%
  rename(dist = spacelag) %>%
  mutate_at("timelag", factor)

variog_empir %>%
  ggplot(aes(dist, gamma, color = timelag, size = n)) +
  # geom_hline(data = variog_sills, aes(yintercept = sills), linetype = "dashed", color = "gray50") +
  # geom_vline(data = variog_data, aes(xintercept = range), linetype = "dashed", color = "gray50") +
  # geom_jitter(shape = 46, alpha = 0.5) +
  # geom_line(size = 1.5, alpha = 0.5) +
  geom_point(alpha = 0.5) +
  # geom_smooth(data = variog_points, aes(fill = timelag), color = FALSE, alpha = 0.5) +
  geom_line(data = filter(variog_fit, timelag == 1), size = 0.5, color = "darkcyan", alpha = 0.8) +
  geom_line(data = filter(variog_fit, timelag == 0), size = 0.5, color = "red1", alpha = 0.8) +
  scale_x_continuous(expand = expand_scale(0, 0), limits = c(0, 25), name = "Distance (m)") +
  scale_color_discrete() +
  scale_y_continuous(name = "Community dissimilarity") +
  facet_grid(metric + amplicon ~ guild, scales = "free")
